# Custom Functions {#sec-functions}

```{r, include = FALSE}
source("R/setup.R")
```

Learn to write custom functions and include them in your package.

```{r}
# packages required for this chapter

library(usethis)  # for setting up data sets in a package
library(glue)     # for text joining
library(devtools) # for loading your package
```

## Function Overview

First, let's create a very basic `r glossary("function")` to learn about custom functions. Functions need a name (like any R `r glossary("object")`). They are created with the `function()` function.

```{r}
fullname <- function() {
  # function code goes here
}
```

Most functions have `r glossary("argument", "arguments")` that set inputs to the function or options for how the function can work. These arguments can be required for the function to work, or have `r glossary("default value", "default values")`. In the example below, the argument `reverse` defaults to FALSE unless you change it.

```{r}
fullname <- function(first, last, reverse = FALSE) {
  # function code goes here
}
```

Functions use these arguments in their code to produce some kind of output (or side effect). Here, if the value of `reverse` is TRUE, then we paste together the value of `last`, a comma, and the value of `first` and assign this to the object `name`. If the value of `reverse` is FALSE, then we paste together the value of `first` and the value of `last` and assign this to `name`. Finally, we use the `return()` function to return the value of `name`.

```{r}
fullname <- function(first, last, reverse = FALSE) {
  if (reverse) {
    name <- paste0(last, ", ", first)
  } else {
    name <- paste(first, last)
  }
  
  return(name)
}
```

:::{.info}
You technically don't have to use the `return()` function. The last object created in the function code will be automatically returned. Most people don't use `return()`, but that can sometimes make it hard to figure out exactly what is being returned if you have a lot of if/else logic.

The `return()` function also stops all subsequent code from being run.
:::

Run the code above to define the function. After it's defined, if you type the function name into the console, without parentheses, it will show you the code for the function.

```{r}
fullname
```

:::{.info}
You can do this for any function; try a few! Many of the base R functions, like `mean` have unsatisfying code like `UseMethod("mean")`, which you can lean about in the [S3 Chapter of Advanced R](https://adv-r.hadley.nz/s3.html), but other functions like `sd` will show you the code they use.
:::

Use your function like any other R function.

```{r}
fullname("Lisa", "DeBruine")
fullname("Lisa", "DeBruine", reverse = TRUE)
```



## Function Development

For `demopkg`, we're going to create a function that produces the APA-formatted text for the results of a paired-samples t-test. Here's an example of APA format.

> A paired-samples t-test was conducted to compare {dv} between {level1} (M = {mean1}, SD = {sd1}) and {level2} (M = {mean2}, SD = {sd2}). There was a {non}significant difference; t({df}) = {t_value}, p = {p_value}.

### Specific Instance

The first step is to sort out a specific instance of your code. We'll load in the data we added last week and compare preferences for self-resembling female faces (`f_self`) to others' preferences for those same faces `f_non`.

```{r}
data("self_res_att", package = "demopkg")

t_results <- t.test(
  x = self_res_att$f_self, 
  y = self_res_att$f_non,
  paired = TRUE)

t_results
```

Now we set up the text template with variables where we want to insert values from the analysis.

```{r}
template <- "A paired-samples t-test was conducted to compare {dv} between {level1} (M = {mean1}, SD = {sd1}) and {level2} (M = {mean2}, SD = {sd2}). There was a {non}significant difference; t({df}) = {t_value}, p = {p_value}."
```

The object `t_results` prints out like above, but the object is actually a list, Use the `str()` function to see the structure of the list.

```{r}
str(t_results)
```

Now we can set the value of each variable from the `t_results` object. You can't get the means and standard deviations from the `t_results` object, so we'll calculate those from the data.

```{r}
glue::glue(
  template,
  dv      = "preferences for female faces",
  level1  = "participants who resembled those faces",
  level2  = "non-self participants",
  mean1   = round(mean(self_res_att$f_self), 1), 
  sd1     = round(sd(self_res_att$f_self), 1),
  mean2   = round(mean(self_res_att$f_non), 1),
  sd2     = round(sd(self_res_att$f_non), 1),
  non     = ifelse(t_results$p.value < .05, "", "non-"),
  df      = round(t_results$parameter, 1),
  t_value = round(t_results$statistic, 2),
  p_value = round(t_results$p.value, 3)
)
```

::: {.info}
Don't worry about `p = 0` just now. A further practice exercise is for you to add code to the function to change it to `p < .001` where appropriate.
:::

### Set up Function

Now we're ready to abstract this into a function. The function will need a name. This function will (for now) only give you the APA-style text for a paired-samples t-test, so let's call it `apa_t_pair`.

You can create an R script in the `R` directory called `apa_t_pair.R`, but the code below does this for you.

```{r, eval = FALSE}
usethis::use_r("apa_t_pair")
```

We'll develop our function in this file. To start, set up a blank function definition.

```{r}
apa_t_pair <- function() {
  
}
```

Then add the code above inside the curly brackets.

```{r}
apa_t_pair <- function() {
  t_results <- t.test(self_res_att$f_self, 
                      self_res_att$f_non,
                      paired = TRUE)
  
  template <- "A paired-samples t-test was conducted to compare {dv} between {level1} (M = {mean1}, SD = {sd1}) and {level2} (M = {mean2}, SD = {sd2}). There was a {non}significant difference; t({df}) = {t_value}, p = {p_value}."
  
  glue::glue(
    template,
    dv      = "preferences for female faces",
    level1  = "participants who resembled those faces",
    level2  = "non-self participants",
    mean1   = round(mean(self_res_att$f_self), 1), 
    sd1     = round(sd(self_res_att$f_self), 1),
    mean2   = round(mean(self_res_att$f_non), 1),
    sd2     = round(sd(self_res_att$f_non), 1),
    non     = ifelse(t_results$p.value < .05, "", "non-"),
    df      = round(t_results$parameter, 1),
    t_value = round(t_results$statistic, 2),
    p_value = round(t_results$p.value, 3)
  )
}
```

Run the code above and test that the function works by running it once in the console.

```{r}
# run in the console
apa_t_pair()
```

::: {.warning}
If you get the message: `Error in apa_t_pair() : could not find function "apa_t_pair"`, this means you didn't run the code that created the function. 
:::

### Add Arguments

We probably want this function to work for any pair of vectors we give it, not just the value of `f_self` and `f_non`. So we need to add `r glossary("argument", "arguments")` to the function. Add arguments `x` and `y` to the function and replace `self_res_att$f_self` with `x` and `self_res_att$f_self` with `y` everywhere in the function.

```{r}
apa_t_pair <- function(x, y) {
  t_results <- t.test(x, y, paired = TRUE)
  
  template <- "A paired-samples t-test was conducted to compare {dv} between {level1} (M = {mean1}, SD = {sd1}) and {level2} (M = {mean2}, SD = {sd2}). There was a {non}significant difference; t({df}) = {t_value}, p = {p_value}."
  
  glue::glue(
    template,
    dv      = "preferences for female faces",
    level1  = "participants who resembled those faces",
    level2  = "non-self participants",
    mean1   = round(mean(x), 1), 
    sd1     = round(sd(x), 1),
    mean2   = round(mean(y), 1),
    sd2     = round(sd(y), 1),
    non     = ifelse(t_results$p.value < .05, "", "non-"),
    df      = round(t_results$parameter, 1),
    t_value = round(t_results$statistic,2),
    p_value = round(t_results$p.value, 3)
  )
}
```

Now, if you try to run the function without any arguments, you'll get an error message. This is because there are no default values for `x` and `y`. 

```{r, error = TRUE}
apa_t_pair()
```

This also won't work. 

```{r, error = TRUE}
x <- self_res_att$f_self
y <- self_res_att$f_non
apa_t_pair()
```

This is because the `x` and `y` inside of the function are in a different `r glossary("environment")` to any `x` and `y` outside of the function. This can seem confusing at first, but it's good that you don't need to worry about objects that exist outside of your function.

You can specify the vectors as arguments.

```{r}
# run in the console or another file
apa_t_pair(x = self_res_att$f_self, 
           y = self_res_att$f_non)
```

### Further Arguments

Now we can further customise our function. You probably won't always be comparing "preferences for female faces" between "participants who resembled those faces" and "non-self participants", so let's add three new arguments to the function. We can set generic `r glossary("default value", "default values")` for these new arguments so that you don't have to specify them if the defaults are OK. 

Since the values are defined with the variable names used in the glue template, we don't need to specify those in the `glue()` function anymore.

```{r}
apa_t_pair <- function(x, y, 
                       dv = "the DV", 
                       level1 = "level 1", 
                       level2 = "level 2") {
  t_results <- t.test(x, y, paired = TRUE)
  
  template <- "A paired-samples t-test was conducted to compare {dv} between {level1} (M = {mean1}, SD = {sd1}) and {level2} (M = {mean2}, SD = {sd2}). There was a {non}significant difference; t({df}) = {t_value}, p = {p_value}."
  
  glue::glue(
    template,
    mean1   = round(mean(x), 1), 
    sd1     = round(sd(x), 1),
    mean2   = round(mean(y), 1),
    sd2     = round(sd(y), 1),
    non     = ifelse(t_results$p.value < .05, "", "non-"),
    df      = round(t_results$parameter, 1),
    t_value = round(t_results$statistic,2),
    p_value = round(t_results$p.value, 3)
  )
}
```

Try running the function both with and without the new arguments. 

```{r}
apa_t_pair(x = self_res_att$f_self, 
           y = self_res_att$f_non)
```


```{r}
apa_t_pair(x = self_res_att$f_self, 
           y = self_res_att$f_non,
           dv = "preferences for female faces",
           level1 = "participants who resembled those faces",
           level2 = "non-self participants")
```

::: {.warning}
If the output doesn't change, this usually means that you forgot to run the code that re-defined `apa_t_pair`. 
:::

## Load in package

### Import dependencies

You need to "import" any non-base packages that you use in a function. These are called "dependencies" because your function depends on them. Our function above uses `glue()` from the glue package. The function below is a quick way to add a dependency.

```{r, eval = FALSE}
usethis::use_package("glue")
```

You should see this output: 

```
✔ Setting active project to '/Users/lisad/rproj/demopkg'
✔ Adding 'glue' to Imports field in DESCRIPTION
• Refer to functions with `glue::fun()`
```

This means that you should always use the full form `glue::glue()`, rather than loading a package with the `library()` function and using just the function name `glue()`. 

You can open the `DESCRIPTION` file to see what has changed. Alternatively, you can manually add dependencies to this file under "Imports:".


### Load 

Now restart R and make sure that your environment is clear. Run the following code to load your package. You can also use a keyboard shortcut to run this function (Mac: cmd-shift-L, Windows: ctl-shift-L).

```{r, eval = FALSE}
devtools::load_all(".")
```

The function should now be available.

```{r}
apa_t_pair(x = self_res_att$f_self, 
           y = self_res_att$f_non,
           dv = "preferences for female faces",
           level1 = "participants who resembled those faces",
           level2 = "non-self participants")
```

And it should be easy to adapt for other pairs of values, such as the equivalent analysis for male faces.

```{r}
apa_t_pair(x = self_res_att$m_self, 
           y = self_res_att$m_non,
           dv = "preferences for male faces",
           level1 = "participants who resembled those faces",
           level2 = "non-self participants")
```


## Glossary

```{r, echo = FALSE, results='asis'}
glossary_table(as_kable = FALSE) |> 
  kableExtra::kable(row.names = FALSE, escape = FALSE) |>
  unclass() |> cat()
```


## Further Practice

1. Edit the function to handle p-values < .001.

1. Add 95% confidence intervals to the output.

1. Allow the user to set a custom confidence interval.

1. Create another function to produce the text for a different analysis you're familiar with in R, such as an ANOVA or correlation.
