# Pkgdown Websites {#sec-pkgdown}

Create a package website using pkgdown.

```{r}
library(usethis)
library(pkgdown)
```

## Set up pkgdown

The following function makes some edits to `Rbuildignore` (telling R to ignore these files when building the package) and `.gitignore` (telling git to ignore the docs directory).

```{r, eval = FALSE}
# run in the console
usethis::use_pkgdown()
```

You'll see some output that looks like this. 

::: {.cell-output}
```
✔ Setting active project to '/Users/lisad/rproj/debruine/demopkg'
✔ Adding '^_pkgdown\\.yml$', '^docs$', '^pkgdown$' to '.Rbuildignore'
✔ Adding 'docs' to '.gitignore'
✔ Writing '_pkgdown.yml'
• Modify '_pkgdown.yml'
```
:::

A file called `_pkgdown.yml` will open with the following contents. 

``` yaml
url: ~
template:
  bootstrap: 5
```

While it is not required, I like to make a directory called `pkgdown` and keep all pkgdown-related contents in there.

```{r, eval = FALSE}
# run in the console
dir.create("pkgdown")
file.rename(from = "_pkgdown.yml",
            to = "pkgdown/_pkgdown.yml")
```


## CITATION file
